Boutiques: a flexible framework for automated application integration in computing platforms

### 2017-11-03 boutiquespaper.tex
Fixed a few TODOs, added Fig 1

\affil[1]{Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada}
\affil[2]{McGill University, Montreal, Canada}
\affil[3]{Montreal Neurological Institute, Montreal, Canada}
\affil[4]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montréal, QC, Canada}
\affil[5]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France}
\affil[6]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA}
\affil[7]{\todo{to be added}}
\affil[8]{Department of Psychology, Stanford University, Stanford, California, USA}
\affil[9]{Partners MS Center, Brigham and Women's Hospital, Brookline, Massachusetts, USA}
\affil[10]{Compute Canada}

@@

\begin{abstract}Porting applications to execution platforms such as web
  portals and workflow engines is critical to enable their sharing
  among scientific communities and their efficient execution on
  computing infrastructures. However, application porting remains a
  costly human effort that consists of 1)~installing the application
  on the target infrastructure, 2)~describing the application in a
  format compatible with the execution platform, and 3)~generating
  proper user interfaces. Due to the variety of computing platforms,
  application porting efforts are often replicated several times while
  interoperability would save cost and improve the quality of the ported
  applications. We describe \boutiques, a descriptive framework and
  command line utility that allows for automatic import and exchange of
  applications across platforms. \boutiques adopts a rich, flexible
  application description format which relies on Linux container
  environments Docker and Singularity to install applications
  across platforms in a lightweight manner. \boutiques is currently used
  by three distinct virtual research platforms, and has been used to
  describe dozens of neuroinformatics tools.
\end{abstract}

\begin{keywords}
Keyword1; keyword 2; keyword 3 (Three to ten keywords representing the main content of the article)
\end{keywords}
\end{frontmatter}

@@

\todo{Intro on the relevance of pipeline sharing for reproducible and open science.}

@@

Platforms such as web services, portals, scientific gateways, workflow
engines and virtual research environments commonly integrate
third-party applications to facilitate the processing of large
datasets on high-performance computing resources. Such applications,
however, are often repeatedly and manually ported to various platforms
whereas this porting effort would greatly benefit from being
mutualized. Meanwhile, virtual container systems such as Docker\todo{cite} and
Singularity\todo{cite} have emerged to facilitate the sharing of software and
improve the reproducibility of analyses by defining immutable,
reusable execution environments. Nevertheless, the automatic
integration of containerized applications in platforms require formal
annotations of container images, to describe the available command lines.

\boutiques is a system to share command-line applications across
platforms. In \boutiques, a command line is described using a flexible
template comprising the inputs it requires and the outputs it
produces. Inputs may be inter-dependent and passed in configuration
files or directly on the command line. Such formal descriptions,
simply referred to as \emph{descriptors}, link to a container image
where the given application and all its dependencies are
installed. \boutiques descriptors allow for automatic application
integration in platforms, and advanced validation of input values to
prevent errors. \boutiques also provides a simple parallelization
framework that allows parallel applications to concurrently execute
multiple tasks (in the remainder, tasks are called
\emph{invocations}). \boutiques descriptors are intented to be
produced by scientific application developers, stored alongside their
application, indexed by common repositories, and consumed by
execution platforms.  A set of core tools facilitate the construction,
validation, import, execution, and publishing of \boutiques
descriptors.

@@

Rigorous input validation is an important motivation for
\boutiques. For this purpose, \boutiques relies on an
application-specific JSON schema\todo{cite}, called \emph{invocation schema}, to
specify the input values accepted by an application. Platforms can
rely on invocation schemas to validate inputs using
any JSON validator, without having to develop specific code.

@@

\subsubsection{Pegasus}
%% gk: to what extent does this exist?
\todo{Rafael}

@@

\subsubsection{Anatomical MRI}

\paragraph{FSL} Several tools from the FMRIB Software
library (FSL) were ported to CBRAIN using \boutiques: BET, fsl\_anat, FAST,
FIRST and probtrackx2. Descriptors are available on Neurolinks. \todo{add reference.}

\paragraph{Anima}
The animaN4BiasCorrection tool from the
Anima\footnote{\url{https://github.com/Inria-Visages/Anima-Public/wiki}} project was also made available 
in VIP. The import process was straightforward. \todo{Sorina: reference?}

\subsubsection{Functional MRI}

\paragraph{Niak} The Niak pre-processing pipeline, executed with the Pipeline System
for Octave and Matlab (PSOM)~\cite{bellec2012pipeline}, was integrated
in CBRAIN through \boutiques. The integration uses the parallelization
mechanism described earlier so that even
the invocations processing a single subject can be parallelized. It also
allows CBRAIN to leverage the efficient agent model used in PSOM, as
described in~\cite{GLAT-16}. The integration required some work in
PSOM to facilitate its invocation as a non-interactive command-line
application. The resulting CBRAIN plugin is available at
\url{https://github.com/SIMEXP/cbrain-plugins-psom}. It can
be used in other platforms that support the parallelization
property. Descriptors are available on Neurolinks.

\paragraph{GinFizz}
Ginfizz, a FMRI pipeline allowing to execute multiple 
preprocessing steps and basic analyses such as: (i) realignment (motion registration and correction),
(ii) slice timing (temporal interpolation to correct for the delay of acquisition between slices), 
(iii) coregistration (registering functional and structural data), (iv) normalization (registering 
of functional and anatomical data to a match a template that has to be specified), and (iv)
smoothing (spatial filtering to increase Signal to Noise ratio). The use of Docker containers
(\url{https://hub.docker.com/r/fliiam/ginfizz} based on \url{https://hub.docker.com/r/miykael/nipype_level2}) 
greatly simplifies its deployment.
\todo{Sorina: reference?}

@@

\paragraph{MRTrix3}
A few tools from the MRtrix3\footnote{\url{http://www.mrtrix.org/}}
package for diffusion MRI processing were also made available in VIP,
as well as a script developed at Creatis, which combines MRtrix3 and
FSL tools. \todo{Sorina: replace footnote by citation, if available.}

@@

\paragraph{ODIN} The Odin MRI simulator\footnote{\url{http://od1n.sourceforge.net}} was imported 
in VIP through \boutiques. Since Odin may require important amounts of
computing resources (the common use-case consists in iterating on a
large number of inputs files), Odin is executed by VIP on the EGI grid
resources, which currently do not support Docker. Consequently, the
Odin executable was exported from the Docker image and the Odin
wrapper script was modified accordingly. \todo{Sorina: replace footnote by citation, if available.}

\subsubsection{General-purpose toolboxes}

\paragraph{BIDS apps} BIDS apps, a recent effort for the adoption of the Brain Imaging Data Structure (BIDS)
in common neuroimaging pipelines, require a standardized set of input
and output parameters~\cite{gorgolewski2017bids}. We developed a tool
as part of the \boutiques importer to generate a descriptor for any
such BIDS app. We validated this tool by importing BIDS apps
containing the Statistical Parameter Mapping toolbox
(SPM)~\cite{penny2011statistical} and the ndmg pipeline mentioned
above. Descriptors are available on Neurolinks.

\paragraph{2016 MICCAI challenges}
We used \boutiques to integrate 23 pipelines in the VIP platform in
the context of two challenges organized by the MICCAI conference in
2016, related to the segmentation of multiple-sclerosis lesions in MR
images (MSSEG
challenge\footnote{\url{https://portal.fli-iam.irisa.fr/msseg-challenge/overview}})
and of tumor volumes in PET images (PETSEG
challenge\footnote{\url{https://portal.fli-iam.irisa.fr/petseg-challenge/overview}}). The
pipelines were ported to VIP and executed on 205 subjects in a few
weeks only, which would not have been possible without
\boutiques. Some pipelines had to be adjusted manually once ported to
the platform for the following reasons:
\begin{enumerate}
\item A pipeline
  required a GPU, which we enabled through the
  \texttt{nvidia-docker}\footnote{\url{https://github.com/NVIDIA/nvidia-docker}}
  tool not supported in Boutiques although it could be a possible extension.
\item A pipeline required more than 10 GB of data dependencies (atlas
  data) which exceeded the maximum size allowed for Docker containers
  in our setup. We solved the issue by installing the data in a
  directory of the host server that we mounted in the container. Such
  platform-specific configurations could not possibly be supported by
  \boutiques.
\item A pipeline wrote more than 10 GB of intermediate data in a
  temporary directory of the container located on a 2 GB partition. We
  solved the issue by mounting a host directory in the temporary directory. 
\end{enumerate}
The pipelines did not use much of the expressiveness features offered
by \boutiques since the instructions given to the challengers required
that all the pipeline parameters were hardcoded, to facilitate
execution. Two custom properties (\texttt{vip:miccai-challenger-email}
and \texttt{vip:miccai-challenge-team-name}) were added to the
\boutiques descriptor to help post-process results in the specific
context of MICCAI challenges.
\todo{Cite Mathieu's paper.}

@@

The recent advent of virtual containers requires a new generation of
application description frameworks that are independent from any
programming language and that expose a rich set of properties to
describe command lines, as intended by \boutiques.

>>>>

\affil[1]{Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada}
\affil[2]{McGill University, Montreal, Canada}
\affil[3]{Montreal Neurological Institute, Montreal, Canada}
\affil[4]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montréal, QC, Canada}
\affil[5]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France}
\affil[6]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA}
\affil[7]{Wellcome Trust Centre for Neuroimaging}
\affil[8]{Department of Psychology, Stanford University, Stanford, California, USA}
\affil[9]{Partners MS Center, Brigham and Women's Hospital, Brookline, Massachusetts, USA}
\affil[10]{Compute Canada}

@@

\begin{abstract}Porting applications to execution platforms such as web
  portals and workflow engines is critical to enable their sharing
  among scientific communities and their efficient execution on
  computing infrastructures. However, application porting remains a
  costly human effort that consists of 1)~installing the application
  on the target infrastructure, 2)~describing the application in a
  format compatible with the execution platform, and 3)~generating
  proper user interfaces. Due to the variety of computing platforms,
  application porting efforts are often replicated several times while
  interoperability would save cost and improve the quality of the ported
  applications. We present \boutiques, a descriptive framework and
  command line utility that allows for automatic import and exchange of
  applications across platforms. \boutiques adopts a rich, flexible
  application description format that relies on Linux container
  environments Docker and Singularity to install applications
  across platforms in a lightweight manner. \boutiques is currently used
  by several distinct virtual research platforms, and has been used to
  describe dozens of neuroinformatics tools.
\end{abstract}

\begin{keywords}
Application integration; Containers; Neuroinformatics.
\end{keywords}
\end{frontmatter}

@@

@@

Platforms such as web services, portals, scientific gateways, workflow
engines and virtual research environments commonly integrate
third-party applications to enable data processing on various types of
computing resources. Such applications, however, are often repeatedly
and manually ported to different platforms whereas sharing this
porting effort would contribute to Open Science and improve
computational reproducibility. Meanwhile, software container systems
such as Docker\footnote{\url{https://www.docker.com}} and
Singularity~\cite{kurtzer2017singularity} have emerged to facilitate
the sharing and migration of software by defining immutable, reusable
execution environments.

We present \boutiques, a system to publish, integrate and execute
containerized command-line applications across platforms (see
Figure~\ref{fig:diagram}).  In \boutiques, a command line is described
using a flexible template comprising the inputs it requires and the
outputs it produces. Inputs may be inter-dependent and passed in
configuration files or directly on the command line. Such formal
descriptions, simply referred to as \emph{descriptors}, link to a
container image where the given application and all its dependencies
are installed. \boutiques descriptors allow for automatic application
integration in platforms, and advanced validation of input values to
prevent errors. \boutiques also provides a simple parallelization
framework that allows parallel applications to concurrently execute
multiple tasks (in the remainder, tasks are called
\emph{invocations}). \boutiques descriptors are intented to be
produced by scientific application developers, stored alongside their
application, indexed by common repositories, and consumed by execution
platforms.  A set of core tools facilitate the construction,
validation, import, execution, and publishing of \boutiques
descriptors.
\begin{figure}
  \includegraphics[width=\columnwidth]{boutiquesfig1.pdf}
  \caption{Publication, integration and execution of applications with \boutiques.}
  \label{fig:diagram}
\end{figure}

@@

Rigorous input validation is an important motivation for
\boutiques. For this purpose, \boutiques relies on an
application-specific JSON schema, called \emph{invocation schema}, to
specify the input values accepted by an application. Platforms can
rely on invocation schemas to validate inputs using
any JSON validator, without having to develop specific code.

@@

@@

\subsubsection{Anatomical imaging}

\paragraph{FSL} Several MRI analysis tools from the FMRIB Software
library (FSL~\cite{jenkinson2012fsl}) were ported to CBRAIN using
\boutiques: BET, fsl\_anat, FAST and FIRST. Descriptors are available
on Neurolinks.

\paragraph{Anima}
\texttt{animaN4BiasCorrection}, an ITK-based bias field correction tool from the
Anima\footnote{\url{https://github.com/Inria-Visages/Anima-Public/wiki}}
project was also made available in VIP. The import process was
straightforward.

\subsubsection{Functional MRI (fMRI)}

\paragraph{Niak} The Niak fMRI pre-processing pipeline, executed with the Pipeline System
for Octave and Matlab (PSOM)~\cite{bellec2012pipeline}, was integrated
in CBRAIN through \boutiques. The integration uses the parallelization
mechanism described earlier so that even the invocations processing a
single subject can be parallelized. It also allows CBRAIN to leverage
the efficient agent model used in PSOM, as described
in~\cite{GLAT-16}. The integration required some work in PSOM to
facilitate its invocation as a non-interactive command-line
application. The resulting CBRAIN plugin is available at
\url{https://github.com/SIMEXP/cbrain-plugins-psom}. It can be used in
other platforms that support the parallelization property. Descriptors
are available on Neurolinks.

\paragraph{GinFizz}
We integrated the
GinFizz\footnote{\url{https://github.com/thomashirsch/ginfizz}}
Nipype-based fMRI pre-processing pipeline in VIP with \boutiques. \todo{Sorina, Pascal, a few lines on the technical deployment?}
%% Ginfizz performs multiple
%% preprocessing steps and basic analyses such as: (i) realignment
%% (motion registration and correction), (ii) slice timing (temporal
%% interpolation to correct for the delay of acquisition between slices),
%% (iii) coregistration (registering functional and structural data),
%% (iv) normalization (registering of functional and anatomical data to a
%% match a template that has to be specified), and (iv) smoothing
%% (spatial filtering to increase Signal to Noise ratio).

@@

\paragraph{MRTrix3}
A few tools from the MRtrix3
package~\cite{tournier2012mrtrix}\footnote{\url{http://www.mrtrix.org/}} for diffusion MRI processing were also made available in VIP,
as well as a script developed at Creatis, which combines MRtrix3 and
FSL tools.

@@

\paragraph{ODIN} The Odin MRI simulator~\cite{jochimsen2004odin}\footnote{\url{http://od1n.sourceforge.net}} was imported 
in VIP through \boutiques. Since Odin may require important amounts of
computing resources (the common use-case consists in iterating on a
large number of inputs files), Odin is executed by VIP on the EGI grid
resources, which currently do not support Docker. Consequently, the
Odin executable was exported from the Docker image and the Odin
wrapper script was modified accordingly.

\subsubsection{BIDS apps}

BIDS apps, an effort for the adoption of the Brain Imaging Data Structure (BIDS)
in common neuroimaging pipelines, require a standardized set of input
and output parameters~\cite{gorgolewski2017bids}. We developed a tool
as part of the \boutiques importer to generate a descriptor for any
such BIDS app. We validated this tool by importing BIDS apps
containing the Statistical Parameter Mapping toolbox
(SPM)~\cite{penny2011statistical} and the ndmg pipeline mentioned
above. Descriptors are available on Neurolinks.

\subsubsection{2016 MICCAI challenges}

We used \boutiques to integrate 23 pipelines in the VIP platform in
the context of two challenges organized by the MICCAI conference in
2016, related to the segmentation of multiple-sclerosis lesions in MR
images (MSSEG
challenge\footnote{\url{https://portal.fli-iam.irisa.fr/msseg-challenge/overview}})
and of tumor volumes in PET images (PETSEG
challenge~\cite{hatt2018}). The
pipelines were ported to VIP and executed on 205 subjects in a few
weeks only, which would not have been possible without
\boutiques. Some pipelines had to be adjusted manually once ported to
the platform for the following reasons:
\begin{enumerate}
\item A pipeline
  required a GPU, which we enabled through the
  \texttt{nvidia-docker}\footnote{\url{https://github.com/NVIDIA/nvidia-docker}}
  tool not supported in Boutiques although it could be a possible extension.
\item A pipeline required more than 10 GB of data dependencies (atlas
  data) which exceeded the maximum size allowed for Docker containers
  in our setup. We solved the issue by installing the data in a
  directory of the host server that we mounted in the container. Such
  platform-specific configurations could not possibly be supported by
  \boutiques.
\item A pipeline wrote more than 10 GB of intermediate data in a
  temporary directory of the container located on a 2 GB partition. We
  solved the issue by mounting a host directory in the temporary directory. 
\end{enumerate}
The pipelines did not use much of the expressiveness features offered
by \boutiques since the instructions given to the challengers required
that all the pipeline parameters were hardcoded, to facilitate
execution. Two custom properties (\texttt{vip:miccai-challenger-email}
and \texttt{vip:miccai-challenge-team-name}) were added to the
\boutiques descriptor to help post-process results in the specific
context of MICCAI challenges.

@@

The recent advent of software containers requires a new generation of
application description frameworks that are independent from any
programming language and that expose a rich set of properties to
describe command lines, as intended by \boutiques.
